# Introductory Example

{{< include _r_setup.qmd >}}

<!-- {{< include _python_setup.qmd >}} -->


## Discrete first


## Sim

Choose a good scenario that highlights prior vs posterior, shrinkage, etc

```{r}
n_rep = 100000
n1 = 5
n2 = 10

sim = data.frame(rep = 1:n_rep) |>
  mutate(p = rbeta(n_rep, 20, 20),
         p1 = pmax(0, pmin(1, rnorm(n_rep, p, 0.1))),
         p2 = pmax(0, pmin(1, rnorm(n_rep, p, 0.1))),
         y1 = rbinom(n_rep, n1, p1),
         y2 = rbinom(n_rep, n2, p2))

sim |> head(10) |> kbl() |> kable_styling()
```

```{r}
n_rep = 100000
n1 = 5
n2 = 10

kappa = 20

sim = data.frame(rep = 1:n_rep) |>
  mutate(p = rbeta(n_rep, 1, 1),
         p1 = rbeta(n_rep, shape1 = p * kappa, shape2 = (1 - p) * kappa),
         p2 = rbeta(n_rep, shape1 = p * kappa, shape2 = (1 - p) * kappa),
         y1 = rbinom(n_rep, n1, p1),
         y2 = rbinom(n_rep, n2, p2))

sim |> head(10) |> kbl() |> kable_styling()
```


## Prior of (p, p1, p2): Scatterplot of p1, p2, colored by p

```{r}
ggplot(sim,
       aes(x = p1,
           y = p2,
           color = p)) +
  geom_point(alpha  = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))
```


```{r}
#| layout-ncol: 2

ggplot(sim,
       aes(x = p1,
           y = p2,
           color = p)) +
  geom_point(alpha  = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))

ggplot(sim,
       aes(x = p,
           y = p2,
           color = p)) +
  geom_point(alpha  = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))

ggplot(sim,
       aes(x = p1,
           y = p,
           color = p)) +
  geom_point(alpha  = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))


```




```{r}
#| layout-ncol: 2

ggplot(sim,
       aes(x = p1,
           y = p2,
           color = p)) +
  geom_point(alpha  = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))

ggplot(sim,
       aes(x = p,
           y = p2,
           color = p1)) +
  geom_point(alpha  = 0.2) +
  scale_color_distiller(palette = "Blues", direction = 1, limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))

ggplot(sim,
       aes(x = p1,
           y = p,
           color = p2)) +
  geom_point(alpha  = 0.2) +
  scale_color_distiller(palette = "Purples", direction = 1, limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))


```



```{r}
sim_hist <- sim |>
  select(p, p1, p2) |>
  pivot_longer(everything(), names_to = "parameter")
```


```{r}
ggplot(sim_hist) +
    stat_density(aes(x = value, color = parameter),
                     geom = "line", position = "identity") +
  scale_color_manual(values = c("seagreen", "skyblue", "purple"))
```




## Fix a few p and look at (p1, p2) to see conditional independence

Fix color so that for given p all dots have same color

```{r}
#| layout-ncol: 3


ggplot(sim |>
         filter(abs(p - 0.3) < 0.05),
       aes(x = p1,
           y = p2,
           color = p)) +
  geom_point(alpha = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  labs(title = "p = 0.3") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))

ggplot(sim |>
         filter(abs(p - 0.5) < 0.05),
       aes(x = p1,
           y = p2,
           color = p)) +
  geom_point(alpha = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  labs(title = "p = 0.5") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))

ggplot(sim |>
         filter(abs(p - 0.7) < 0.05),
       aes(x = p1,
           y = p2,
           color = p)) +
  geom_point(alpha = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  labs(title = "p = 0.7") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))
```

## Scatterplot of y1, p1, colored by p

Arrange 2x2

```{r}
#| layout-ncol: 2

ggplot(sim,
       aes(x = p1,
           y = y1,
           color = p)) +
  geom_jitter(width = 0, height = 0.2, shape = 21, alpha = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  labs(title = "p1 and y1") +
  scale_x_continuous(limits = c(0, 1))

ggplot(sim,
       aes(x = p2,
           y = y2,
           color = p)) +
  geom_jitter(width = 0, height = 0.2, shape = 21, alpha = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  labs(title = "p2 and y2") +
  scale_x_continuous(limits = c(0, 1))

ggplot(sim,
       aes(x = p1,
           y = p,
           color = p)) +
  geom_point() +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  labs(title = "p1 and p") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))

ggplot(sim,
       aes(x = p2,
           y = p,
           color = p)) +
  geom_point() +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  labs(title = "p2 and p") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))
```

## Condition on observed data

Only keep pairs with observed y1, y2 then redo above scatterplots with all other pairs removed

First show the (p1, y1) scatterplot with row highlighted and then show that corresponding points in the (p1, p) plot with the same highlight color

```{r}

y1_obs = 1
y2_obs = 7

sim_post <- sim |>
  filter(y1 == y1_obs,
         y2 == y2_obs)

sim_post |> head(10) |> kbl() |> kable_styling()
```


Need to keep only values with both y1 and y2

```{r}
#| layout-ncol: 2


ggplot(sim,
       aes(x = p1,
           y = y1,
           color = p)) +
  geom_jitter(width = 0, height = 0.2, shape = 21, alpha = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  annotate("rect", xmin = 0, xmax = 1,
           ymin = y1_obs - 0.4, ymax = y1_obs + 0.4, alpha = 0.5,
           color = "orange",
           fill = "orange") +
  labs(title = "p1 and y1, given y1 obs")

ggplot(sim,
       aes(x = p2,
           y = y2,
           color = p)) +
  geom_jitter(width = 0, height = 0.2, shape = 21, alpha = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  annotate("rect", xmin = 0, xmax = 1,
           ymin = y2_obs - 0.4, ymax = y2_obs + 0.4, alpha = 0.5,
           color = "skyblue",
           fill = "skyblue") +
  labs(title = "p2 and y2, given y2 obs")


ggplot(sim |>
         mutate(is_y1 = (y1 == y1_obs)),
       aes(x = p1,
           y = p,
           color = is_y1)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("seagreen", "orange")) +
  labs(title = "p1 and p, given y1 obs")

ggplot(sim |>
         mutate(is_y2 = (y2 == y2_obs)),
       aes(x = p2,
           y = p,
           color = is_y2)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("seagreen", "orange")) +
  labs(title = "p2 and p, given y2 obs")
```


```{r}
#| layout-ncol: 2


ggplot(sim,
       aes(x = p1,
           y = y1,
           color = p)) +
  geom_jitter(width = 0, height = 0.2, shape = 21, alpha = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  annotate("rect", xmin = 0, xmax = 1,
           ymin = y1_obs - 0.4, ymax = y1_obs + 0.4, alpha = 0.5,
           color = "orange",
           fill = "orange") +
  labs(title = "p1 and y1, given y1 obs")

ggplot(sim,
       aes(x = p2,
           y = y2,
           color = p)) +
  geom_jitter(width = 0, height = 0.2, shape = 21, alpha = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  annotate("rect", xmin = 0, xmax = 1,
           ymin = y2_obs - 0.4, ymax = y2_obs + 0.4, alpha = 0.5,
           color = "orange",
           fill = "orange") +
  labs(title = "p2 and y2, given y2 obs")

ggplot(sim_post,
       aes(x = p1,
           y = p,
           color = p)) +
  geom_point() +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  labs(title = "p2 and y2")

ggplot(sim_post,
       aes(x = p2,
           y = p,
           color = p)) +
  geom_point() +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  labs(title = "p2 and p")
```





```{r}
#| layout-ncol: 3


ggplot(sim,
       aes(x = y1,
           y = y2,
           color = p)) +
  geom_jitter(width = 0.2, height = 0.2, shape = 21, alpha = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  annotate("rect", ymin = -Inf, ymax = Inf,
           xmin = y1_obs - 0.4, xmax = y1_obs + 0.4, alpha = 0.5,
           color = "orange",
           fill = "orange") +
    annotate("rect", xmin = -Inf, xmax = Inf,
           ymin = y2_obs - 0.4, ymax = y2_obs + 0.4, alpha = 0.5,
           color = "orange",
           fill = "orange") +
  labs(title = "p given observed y1 and y2")

ggplot(sim,
       aes(x = y1,
           y = y2,
           color = p1)) +
  geom_jitter(width = 0.2, height = 0.2, shape = 21, alpha = 0.2) +
  scale_color_distiller(palette = "Blues", direction = 1, limits = c(0, 1)) +
  annotate("rect", ymin = -Inf, ymax = Inf,
           xmin = y1_obs - 0.4, xmax = y1_obs + 0.4, alpha = 0.5,
           color = "orange",
           fill = "orange") +
    annotate("rect", xmin = -Inf, xmax = Inf,
           ymin = y2_obs - 0.4, ymax = y2_obs + 0.4, alpha = 0.5,
           color = "orange",
           fill = "orange") +
  labs(title = "p1 given observed y1 and y2")

ggplot(sim,
       aes(x = y1,
           y = y2,
           color = p2)) +
  geom_jitter(width = 0.2, height = 0.2, shape = 21, alpha = 0.2) +
  scale_color_distiller(palette = "Purples", direction = 1, limits = c(0, 1)) +
  annotate("rect", ymin = -Inf, ymax = Inf,
           xmin = y1_obs - 0.4, xmax = y1_obs + 0.4, alpha = 0.5,
           color = "orange",
           fill = "orange") +
    annotate("rect", xmin = -Inf, xmax = Inf,
           ymin = y2_obs - 0.4, ymax = y2_obs + 0.4, alpha = 0.5,
           color = "orange",
           fill = "orange") +
  labs(title = "p2 given observed y1 and y2")

```



## Posterior of (p, p1, p2): Scatterplot of p1, p2, colored by p

```{r}
ggplot(sim_post,
       aes(x = p1,
           y = p2,
           color = p)) +
  geom_point(alpha  = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))
```


```{r}
#| layout-ncol: 2

ggplot(sim_post,
       aes(x = p1,
           y = p2,
           color = p)) +
  geom_point(alpha  = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))

ggplot(sim_post,
       aes(x = p,
           y = p2,
           color = p)) +
  geom_point(alpha  = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))

ggplot(sim_post,
       aes(x = p1,
           y = p,
           color = p)) +
  geom_point(alpha  = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))


```




```{r}
#| layout-ncol: 2

ggplot(sim_post,
       aes(x = p1,
           y = p2,
           color = p)) +
  geom_point(alpha  = 0.2) +
  scale_color_distiller(palette = "Greens", direction = 1, limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))

ggplot(sim_post,
       aes(x = p,
           y = p2,
           color = p1)) +
  geom_point(alpha  = 0.2) +
  scale_color_distiller(palette = "Blues", direction = 1, limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))

ggplot(sim_post,
       aes(x = p1,
           y = p,
           color = p2)) +
  geom_point(alpha  = 0.2) +
  scale_color_distiller(palette = "Purples", direction = 1, limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))


```


```{r}
sim_post_hist <- sim_post |>
  select(p, p1, p2) |>
  pivot_longer(everything(), names_to = "parameter")
```


```{r}
ggplot(sim_post_hist) +
    stat_density(aes(x = value, color = parameter),
                     geom = "line", position = "identity") +
  scale_color_manual(values = c("seagreen", "skyblue", "purple"))
```

