# Introduction to Bayesian Hierarchical Models {#sec-hierarchical}

{{< include _r_setup.qmd >}}

{{< include _python_setup.qmd >}}



In a **Bayesian hierarchical^[Hierarchical models are often called *multilevel, mixed effects, or random effects models*.] model**, the model for the data depends on certain parameters, and those parameters in turn depend on other parameters.
Hierarchical modeling provides a framework for building
complex and high-dimensional models from simple and
low-dimensional building blocks.


Suppose we want to estimate the true, long run free throw percentage for each basketball player currently on the Philadelphia 76ers^[Along with one bonus "player" who will illustrate some ideas.].
That is, we want to estimate each player's true probability of successfully scoring on any free throw attempt.
We'll consider data from games in the 2022-2023 NBA season obtained from [basketball-reference](https://www.basketball-reference.com/teams/PHI/2023.html#totals::17) as of Feb 27, summarized in @sec-hierarchical-sample.

```{r}
#| include: false
#| echo: false
#| warning: false

data = read_csv("sixers.csv") |>
  rename(FTpct = "FT%")

```



::: {#exm-hier-intro}
Let $\theta_j$ denote player $j$'s true (long run) FT percentage; for example, $\theta_{\text{Embiid}}$, $\theta_{\text{Harden}}$, etc.
Let $n_j$ denote the number of free throw attempts and $y_j$ the number of free throws made for player $j$.
:::

1. One way to estimate $\theta_j$ is to pool all the data together.
Over the `r nrow(data)` players in the data set, there are `r sum(data$FTA)` attempts, of which `r sum(data$FT)` were successful, for an overall free throw percentage of `r 100 * round(sum(data$FT) / sum(data$FTA), 3)`%.  The sample mean free throw percentage, among players with nonzero attempts, is `r 100 * round(mean(data$FTpct, na.rm = TRUE), 3)`%.
So we could just estimate each $\theta_j$ with the overall free throw percentage (or the sample mean free throw percentage).
What are some issues with this approach?
\
\
\
\
\

1. At the other extreme, we could estimate $\theta_j$ using just the attempts for player $j$.
For example, use only the attempts for Embiid to estimate $\theta_{\text{Embiid}}$, only the attempts for Harden to estimate $\theta_{\text{Harden}}$, etc.
What are some issues with this approach?
(Hint: consider the players with no attempts, or few attempts.)
\
\
\
\
\

1. In a Bayesian model, the $\theta_j$'s would have a prior distribution.
We could assume a Bayesian model with a prior in which all $\theta_j$ are independent.
What are some issues with this approach?
\
\
\
\
\

::: {#exm-hier-hyper}
Let's first consider just a single player with true (long run) free throw percentage $\theta$.
Given $n$ independent attempts, conditional on $\theta$, we might assume that $y$ follows a Binomial($n, \theta$) distribution.
:::

1. If we imagine the player is randomly selected from among current NBA players, how might we interpret the prior distribution for $\theta$?
\
\
\
\
\

1. The prior distribution for $\theta$ has a prior mean; let's denote it $\mu$.
How might we interpret $\mu$?
\
\
\
\
\

1. The prior mean $\mu$ is also a parameter, so we could place a prior distribution on it.
What would the prior distribution on $\mu$ represent?
\
\
\
\
\

1. In this model $\mu$ and $\theta$ are dependent.
Why is this dependence useful?
\
\
\
\
\

1. The prior distribution for $\theta$ has a prior standard deviation; let's denote it $\sigma$.
How might we interpret $\sigma$?
\
\
\
\
\

1. The prior standard deviation $\sigma$ is also a parameter, so we could place a prior distribution on it.
What would the prior distribution on $\sigma$ represent?
\
\
\
\
\


- A Bayesian hierarchical model consists of the following layers in the hierarchy.
    1. A model for the data $y$ which depends on model parameters $\theta$ and is represented by the likelihood $f(y|\theta)$.
        - The model usually involves multiple parameters.
        Generically, $\theta$ represents a vector of parameters that show up in the likelihood function.
        - Remember: While the likelihood for a particular $y$ sample is computed by evaulating the conditional density of $y$ given $\theta$, the likelihood function is *a function of $\theta$, treating the data $y$ as fixed*.
    1. A (joint) prior distribution for the model parameter $\theta$, denoted $\pi(\theta|\phi)$, which depends on *hyperparameters* $\phi$.
    1. A *hyperprior* distribution, i.e., a prior distribution for the hyperparameters, denoted $\pi(\phi)$.
(Again, $\phi$ generically represents a vector of hyperparameters.)
- The above specification implicitly assumes that the data distribution and likelihood function only depend on $\theta$; the hyperparameters only affect $y$ through $\theta$.
That is, the general hierarchical model above assumes that the data $y$ and hyperparameters $\phi$ are conditionally independent given the model parameters $\theta$, so the likelihood satisfies
$$
f(y|\theta, \phi) = f(y|\theta)
$$
The joint prior distribution on parameters and hyperparameters is
$$
\pi(\theta, \phi) = \pi(\theta|\phi)\pi(\phi)
$$
- Given the observed data $y$, we find the joint posterior density of the parameters and hyperparameters, denoted $\pi(\theta, \phi|y)$, using Bayes rule: posterior $\propto$ likelihood $\times$ prior.
\begin{align*}
\pi(\theta, \phi|y) & \propto f(y|\theta, \phi)\pi(\theta, \phi)\\
& \propto f(y|\theta)\pi(\theta|\phi)\pi(\phi)
\end{align*}
- In Bayesian hierarchical models, there will generally be dependence between parameters.
But dependence among parameters is actually useful for several reasons:
    - The dependence is natural and meaningful in the problem context.
    - If parameters are dependent, then all data can inform parameter estimates.


::: {#exm-hier-model}
Now suppose want to estimate the free throw percentages for $J$ NBA players, given data consisting of the number of free throws attempted $n_j$ and made $y_j$ by player $j$.
Specify a hierarchical model, including your choice of prior distribution.
How many parameters are there in the model?
:::
\
\
\
\
\


::: {#exm-hier-hyper-post}
Use software to fit the Bayesian model using the 76ers data.
Summarize the posterior distribution of $\mu$.
:::
\
\
\
\
\


::: {#exm-hier-post}
For each player $j$, consider the posterior mean and posterior SD of $\theta_j$.
:::


1. What is the posterior mean of $\theta_j$ for players with no attempts?
Why does this make sense?
\
\
\
\
\

1. Why is the posterior mean of $\theta_j$ not the same for players with no attempts?
\
\
\
\
\


1. For which players is the posterior mean closest to the observed free throw percentage?
Why?
\
\
\
\
\

1. Which players have the smallest posterior SDs?
Why?
\
\
\
\
\

1. Find and interpret 50%, 80%, and 98% posterior credible intervals for Joel Embiid.
\
\
\
\
\

1. Find and interpret 50%, 80%, and 98% posterior credible intervals for Kevin Ross.
\
\
\
\
\





## Notes


### Sample data {#sec-hierarchical-sample}

```{r}
#| warning: false

data = read_csv("sixers.csv") |>
  rename(FTpct = "FT%") |>
  arrange(desc(FTA)) 
```


```{r}
data |>
  select(Player, FTA, FT, FTpct) |>
  kbl() |> kable_styling()
```



### Approximating the posterior distribution

`brms` is structured around linear models.
In the situation of this example, `brms` wants to fit a logistic regression model.
Rather than doing that, I used another software program, JAGS, to fit our current model.
Don't worry about the syntax.
Just notice that the inputs are: (1) the data, (2) the model for the data which determines the likelihood (`dbinom`), and (3) the prior (`dbeta`) and hyperprior (`dnorm`).
And the output is an approximation of the posterior distribution of the parameters.



```{r}

library(rjags)

player = data$Player
y = data$FT
n = data$FTA
J = length(y)

# specify the model: likelihood, prior, hyperprior

model_string <- "model{

  # Likelihood
  for (j in 1:J){
    y[j] ~ dbinom(theta[j], n[j])
  }

  # Prior Beta(alpha, beta)
  # We put hyperpriors on the mean mu and the SD sigma,
  # So we need to convert to shape parameters alpha and beta
  
  for (j in 1:J){
    theta[j] ~ dbeta(mu * (mu * (1 - mu) / sigma ^ 2 - 1), (1 - mu) * (mu * (1 - mu) / sigma ^ 2 - 1))
  }
  
  # Hyperprior, phi=(mu, sigma)
  # In JAGS the second parameter of dnorm is precision = 1 / SD ^ 2
  
  mu ~ dnorm(0.75, 1 / 0.05 ^ 2)
  sigma ~ dnorm(0.08, 1 / 0.02 ^ 2)
}"

dataList = list(y = y, n = n, J = J)

model <- jags.model(file = textConnection(model_string), 
                    data = dataList)

update(model, n.iter = 1000)

posterior_sample <- coda.samples(model,
                       variable.names = c("theta", "mu", "sigma"),
                       n.iter = 10000,
                       n.chains = 5)
```


```{r}
head(posterior_sample)

```




```{r}
summary(posterior_sample)
```


### Posterior distribution of $\mu$

```{r}
mu_post = as.matrix(posterior_sample)[, "mu"]

hist(mu_post, breaks = 100)

quantile(mu_post, c(0.01, 0.10, 0.25, 0.5, 0.75, 0.90, 0.99))
```

### Posterior mean and SD of $\theta_j$s

```{r}
player_summary = data.frame(player, n, y, y / n)
names(player_summary) = c("player", "FTA", "FT", "FTpct")

# In posterior_sample list, the first two entries correspond to mu and sigma,
# so we remove them.

player_summary[c("posterior_mean", "posterior_SD")] =
  summary(posterior_sample)$statistics[-(1:2), c("Mean", "SD")]


player_summary |> kbl(digits = 4) |> kable_styling()
```



```{r}
#| warning: false

player_summary |>
  select(player, FTpct, posterior_mean) |>
  pivot_longer(!player, names_to = "source", values_to = "value") |>
  mutate(label = if_else(source == "FTpct", player, "")) |>
  ggplot(aes(x = source,
           y = value,
           group = player,
           col = player,
           label = label)) +
  scale_color_discrete() +
  geom_point() +
  geom_text(hjust = 1, size = 3) + 
  geom_line(alpha = 0.5) +
  theme(legend.position = "none")
```

### Posterior distribution of $\theta_{\text{Embiid}}$

```{r}
mu_post = as.matrix(posterior_sample)[, "theta[1]"]

hist(mu_post, breaks = 100)

quantile(mu_post, c(0.01, 0.10, 0.25, 0.5, 0.75, 0.90, 0.99))
```


### Posterior distribution of $\theta_{\text{Ross}}$

```{r}
mu_post = as.matrix(posterior_sample)[, "theta[19]"]

hist(mu_post, breaks = 100)

quantile(mu_post, c(0.01, 0.10, 0.25, 0.5, 0.75, 0.90, 0.99))
```



[`ShinyStan`](https://mc-stan.org/users/interfaces/shinystan) provides a nice interface to interactively investigate the posterior distribution.

```{r}
#| eval: false

library(shinystan)

sixers_sso <- as.shinystan(posterior_sample, model_name = "76ers Free Throws")
sixers_sso <- drop_parameters(sixers_sso, pars = c("sigma"))
launch_shinystan(sixers_sso)
```



