# Introduction to Bayesian Hierarchical Linear Models {#sec-hierarchical-linear}

{{< include _r_setup.qmd >}}

{{< include _python_setup.qmd >}}


We have studied a Bayesian simple linear regression model.
We have also introduced some basic principles of hierarchical models.
Now we'll combine elements to formulate Bayesian hierarchical linear regression models.


*This handout corresponds to [Chapters 15-17](https://www.bayesrulesbook.com/chapter-15.html) of [@johnson2022].*
*This handout only contains a brief introduction; see [@johnson2022] for a much more thorough analysis.*

We'll consider the `cherry_blossom_sample` data in the `bayesrules` package which contains the `net` running times (in minutes) and `age` for participants in the annual 10-mile Cherry Blossom race held in Washington, D.C.
Each runner in the sample is in their 50s or 60s.

```{r}
library(bayesrules)

data(cherry_blossom_sample)
running <- cherry_blossom_sample |>
  select(runner, age, net) |>
  na.omit()

head(running)
```


1. Suppose we fit a Bayesian simple regression model of `net` on `age`.
What do the slope and intercept parameters represent in context?
What might be a reasonable prior?
(You can assume that the `age` variable has been centered by subtracting the sample mean.)
\
\
\
\
\

1. Fit the model from the previous part (see the results in [Section 17.1](https://www.bayesrulesbook.com/chapter-17.html#first-steps-complete-pooling) of [@johnson2022]).
Do the results seem reasonable?
(Hint: consider the posterior distribution of the slope.)
\
\
\
\
\

1. Each of the runners in the sample has run the race in multiple years.
What assumption of the simple linear regression model is violated?
\
\
\
\
\

1. Formulate a Bayesian hierarchical linear model which assumes the intercept (but not slope) varies by runner.
What do the parameters represent in context?
\
\
\
\
\
\
\
\
\
\



1. Fit the model from the previous part (see the results in [Section 17.2.4](https://www.bayesrulesbook.com/chapter-17.html#chapter-17-random-intercepts) of [@johnson2022]).
What are some features of the analysis you would explore?
Consider the posterior distribution of the slope associated with `age`; does the posterior distribution based on this hierarchical model seem more reasonable than the one based on the simple linear regression model?
\
\
\
\
\
\
\
\
\
\


1. Formulate a Bayesian hierarchical linear model which assumes both the intercept and the slope vary by runner.
What do the parameters represent in context?
\
\
\
\
\
\
\
\
\
\



1. Fit the model from the previous part (see the results in [Section 17.3.3](https://www.bayesrulesbook.com/chapter-17.html#ch-17-rand-int-slope-sim) of [@johnson2022]).
What are some features of the analysis you would explore?
\
\
\
\
\
\
\
\
\
\

1. Compare the two hierarchical regression models.
What are some similarities?
What are some differences?
\
\
\
\
\
\
\
\
\
\



## Notes

```{r}
library(brms)
library(bayesplot)
library(tidybayes)
```

### Linear model


```{r}
fit_lm <- brm(data = running,
              family = gaussian(),
              net ~ 1 + age)
```


```{r}
summary(fit_lm)
```

### Hierarchical model - random intercepts only


```{r}
fit_hier1 <- brm(data = running,
                 family = gaussian(),
                 net ~ 1 + age + (1 | runner))
```


```{r}
summary(fit_hier1)
```




### Hierarchical model - random intercepts and slopes


```{r}
fit_hier2 <- brm(data = running,
                 family = gaussian(),
                 net ~ 1 + age + (1 + age | runner))
```


```{r}
summary(fit_hier2)
```

```{r}
#| eval: false

library(shinystan)

launch_shinystan(fit_hier2)
```


