# Introduction to Multi-Parameter models {#sec-multi-parameter}

{{< include _r_setup.qmd >}}

<!-- {{< include _python_setup.qmd >}} -->







- Most interesting problems involve multiple unknown parameters.
- When there are two (or more) unknown parameters the prior and posterior distribution will each be a *joint* probability distribution over *pairs* (or tuples/vectors) of possible values of the parameters.




::: {#exm-body-temp-discrete-sd}
Assume body temperatures (degrees Fahrenheit) of healthy adults follow a Normal distribution with unknown mean $\mu$ and unknown standard deviation $\sigma$.
Suppose we wish to estimate both $\mu$, the population mean healthy human body temperature, and $\sigma$, the population standard deviation of body temperatures.
Assume a discrete prior distribution according to which

- $\mu$ takes values 97.6, 98.1, 98.6 with prior probability 0.2, 0.3, 0.5, respectively.
- $\sigma$ takes values 0.5, 1 with prior probability 0.25, 0.75, respectively.
- $\mu$ and $\sigma$ are independent.
:::

1. Start to construct the Bayes table.
What are the possible values of the parameter (in this discrete example)?
What are the prior probabilities?
(Hint: the parameter $\theta$ is a *pair* $(\mu, \sigma)$.)
\
\
\
\
\

1. Suppose two temperatures of 97.5 and 97.9 are observed, independently.
Identify the likelihood.
\
\
\
\
\

1. Complete the Bayes table and find the posterior distribution after observing these two measurements.
Compare to the prior distribution.
\
\
\
\
\


1. The prior assumes that $\mu$ and $\sigma$ are independent.
Are they independent according to the posterior distribution?
\
\
\
\
\



::: {#exm-body-temp-prior-sd}
Continuing @exm-body-temp-discrete-sd, let's assume a more reasonable, continuous prior for $(\mu,\sigma)$.
:::


1. What are the possible values of $\sigma$?
How would you choose a prior for $\sigma$?
\
\
\
\
\


1. For a prior, assume that: $\mu$ has a Normal distribution with mean 98.6 and standard deviation 0.3, $\sigma$ has a "Half-Normal(0, 1)" distribution (that's Normal(0, 1) but constrained to positive values), and $\mu$ and $\sigma$ are independent.
Simulate $(\mu, \sigma)$ pairs from the prior distribution and plot them.
\
\
\
\
\

1. Simulate the prior predictive distribution.
Do the values seem reasonable?
If not, adjust the prior distribution of parameters and try again.
\
\
\
\
\



1. Find and interpret a central 98% prior credible interval for $\mu$.
\
\
\
\
\

1. Find and interpret a central 98% prior credible interval for $\sigma$.
\
\
\
\
\

1. What is the prior credibility that both $\mu$ and $\sigma$ lie within their credible intervals?
\
\
\
\
\


::: {#exm-body-temp-grid-sd}
Continuing @exm-body-temp-discrete-sd, we'll now use grid approximation to approximate the posterior distribution.
:::



1. Assume a grid of $\mu$ values from 96 to 100 in increments of 0.01, and a grid of $\sigma$ values from 0.01 to 3 in increments of 0.01.
How many possible values of the pair $(\mu, \sigma)$ are there; that is, how many rows are there in the Bayes table?
\
\
\
\
\

1. Suppose first that we are given a sample of two measurements of 97.9 and 97.5.
Use grid approximation to approximate the joint posterior distribution of $(\mu, \sigma)$
Simulate values from the joint posterior distribution and plot them.
Compare to the prior.
Compute the posterior correlation between $\mu$ and $\sigma$; are they independent according to the posterior distribution?
\
\
\
\
\


1. In a recent study, the sample mean body temperature in a sample of 208 healthy adults was 97.7 degrees F and the sample standard deviation was 1.0 degree F; the sample data is summarized in @sec-body-temperature-data (it's the same data we used in @sec-brms).
Use grid approximation to approximate the joint posterior distribution of $(\mu, \sigma)$
Simulate values from the joint posterior distribution and plot them.
Compare to the prior.
Compute the posterior correlation between $\mu$ and $\sigma$; are they independent according to the posterior distribution?
\
\
\
\
\


::: {#exm-body-temp-brms-sd}
Continuing @exm-body-temp-discrete-sd, we'll now use simulation to approximate the posterior distribution, given the same sample data as before.
:::



1. Describe how, in principle, you would use a naive (non-MCMC) simulation to approximate the posterior distribution.
In practice, what is the problem with such a simulation?
\
\
\
\
\

1. Use `brms` to approximate the posterior distribution.
Compare to the results of the grid approximation.
\
\
\
\
\

1. Find and interpret a central 98% posterior credible interval for $\mu$.
\
\
\
\
\

1. Find and interpret a central 98% posterior credible interval for $\sigma$.
\
\
\
\
\

1. What is the posterior credibility that both $\mu$ and $\sigma$ lie within their credible intervals?
\
\
\
\
\




::: {#exm-body-temp-prediction-sd}

Continuing the previous example
:::


1. Describe how you could use simulation to approximate the posterior predictive distribution.
\
\
\
\
\

1. Run the simulation, summarize the predictive distribution, and find and interpret a 95% posterior predictive interval.
\
\
\
\
\

1. Explain how you could use posterior predictive checking to assess the reasonableness of the model in light of the observed data.
Does the model seem reasonable?
\
\
\
\
\


## Notes

### Discrete grid

```{r}
mu = c(97.6, 98.1, 98.6)
sigma = c(0.5, 1)

# theta is all possible (mu, sigma) pairs
theta = expand.grid(mu, sigma)
names(theta) = c("mu", "sigma")

# prior
prior_mu = c(0.20, 0.30, 0.50)
prior_sigma = c(0.25, 0.75)
prior = apply(expand.grid(prior_mu, prior_sigma), 1, prod)
prior = prior / sum(prior)

# data
y = c(97.9, 97.5) # single observed value

# likelihood
likelihood = dnorm(97.9, mean = theta$mu, sd = theta$sigma) *
  dnorm(97.5, mean = theta$mu, sd = theta$sigma)

# posterior
product = likelihood * prior
posterior = product / sum(product)

# bayes table
bayes_table = data.frame(theta,
                     prior,
                     likelihood,
                     product,
                     posterior)

# display table
bayes_table |>
  adorn_totals("row",,,, -c(mu, sigma)) |>
  kbl(digits = 4) |>
  kable_styling()
```

```{r}
#| label: fig-body-temp-discrete-tile
#| echo: true
#| fig-cap: "Joint distribution of mean and SD of body temperatures in @exm-body-temp-discrete-sd."
#| fig-subcap: 
#|   - "Prior distribution."
#|   - "Posterior distribution."
#| layout-ncol: 2

ggplot(bayes_table |>
         mutate(mu = factor(mu),
                sigma = factor(sigma)),
       aes(mu, sigma)) +
  geom_tile(aes(fill = prior))  +
  scale_fill_viridis(option = "mako",
                     limits = c(0, max(c(prior, posterior)))) +
  theme_bw()

ggplot(bayes_table |>
         mutate(mu = factor(mu),
                sigma = factor(sigma)),
       aes(mu, sigma)) +
  geom_tile(aes(fill = posterior)) +
  scale_fill_viridis(option = "mako",
                     limits = c(0, max(c(prior, posterior))))
```


### Prior distribution

```{r}
#| label: fig-body-temp-prior
#| echo: true
#| fig-cap: "Simulated joint prior distribution of mean and SD of body temperatures in @exm-body-temp-prior-sd."
#| fig-subcap: 
#|   - "Scatterplot."
#|   - "Density plot."
#| layout-ncol: 2

n_rep = 10000

mu_sim_prior = rnorm(n_rep, 98.6, 0.3)
sigma_sim_prior = abs(rnorm(n_rep, 0, 1))
sim_prior = data.frame(mu_sim_prior, sigma_sim_prior)

ggplot(sim_prior,
       aes(x = mu_sim_prior,
           y = sigma_sim_prior)) +
  geom_point(color = bayes_col["prior"], alpha = 0.4) +
  labs(x = "mean (mu)",
       y = "sd (sigma)") +
  theme_bw()

ggplot(sim_prior,
       aes(x = mu_sim_prior,
           y = sigma_sim_prior)) +
  stat_density_2d(aes(fill = after_stat(level)),
                  geom = "polygon", color = "white") +
  scale_fill_viridis(option = "mako") +
  labs(x = "mean (mu)",
       y = "sd (sigma)") +
  theme_bw()
```


### Prior predictive distribution


```{r}

# given theta = (mu, sigma), simulation y from Normal(mu, sigma)
y_sim_prior = rnorm(n_rep, mu_sim_prior, sigma_sim_prior)

# plot the simulated y values
ggplot(data.frame(y_sim_prior),
       aes(x = y_sim_prior)) + 
  geom_histogram(aes(y = after_stat(density)),
                 col = bayes_col["prior_predict"],
                 fill = "white") +
  geom_density(col = bayes_col["prior_predict"],
               linewidth = 1) +
  labs(x = "Body temperature (y)") +
  theme_bw()

# plot the cdf
ggplot(data.frame(y_sim_prior),
       aes(x = y_sim_prior)) + 
  stat_ecdf(col = bayes_col["prior_predict"],
            linewidth = 1) +
  labs(x = "Body temperature (y)",
       y = "cdf") +
  theme_bw()
```

```{r}
quantile(y_sim_prior, c(0.025, 0.975))
```


### Grid approximation: $n=2$


```{r}
mu = seq(96, 100, 0.01)
sigma = seq(0.01, 3, 0.01)

# theta is all possible (mu, sigma) pairs
theta = expand.grid(mu, sigma)
names(theta) = c("mu", "sigma")

# prior
prior_mu = dnorm(mu, 98.6, 0.3)
prior_sigma = dnorm(sigma, 0, 1)
prior = apply(expand.grid(prior_mu, prior_sigma), 1, prod)
prior = prior / sum(prior)

# data
y = c(97.9, 97.5) # single observed value

# likelihood
likelihood = dnorm(97.9, mean = theta$mu, sd = theta$sigma) *
  dnorm(97.5, mean = theta$mu, sd = theta$sigma)

# posterior
product = likelihood * prior
posterior = product / sum(product)

```




```{r}
#| label: fig-body-temp-posterior-discrete
#| echo: true
#| fig-cap: "Grid-approximated joint posterior distribution of mean and SD of body temperatures in @exm-body-temp-prior-sd."
#| fig-subcap: 
#|   - "Scatterplot."
#|   - "Density plot."
#| layout-ncol: 2

n_rep = 10000

# sample (mu, sigma) pairs for the posterior distribution by
# sampling rows of the Bayes table according to posterior
# and then finding the (mu, sigma) values for the sampled rows

index_sim = sample(1:length(posterior),
                   size = n_rep, replace = TRUE,
                   prob = posterior)

sim_posterior = theta |> slice(index_sim)

ggplot(sim_posterior,
       aes(x = mu,
           y = sigma)) +
  geom_point(color = bayes_col["posterior"], alpha = 0.4) +
  labs(x = "mean (mu)",
       y = "sd (sigma)") +
  theme_bw()

ggplot(sim_posterior,
       aes(x = mu,
           y = sigma)) +
  stat_density_2d(aes(fill = after_stat(level)),
                  geom = "polygon", color = "white") +
  scale_fill_viridis(option = "mako") +
  labs(x = "mean (mu)",
       y = "sd (sigma)") +
  theme_bw()
```

```{r}
cor(sim_posterior$mu, sim_posterior$sigma)
```


### Sample data {#sec-body-temperature-data}

```{r}
data = read_csv("bodytemps.csv")

head(data)
```


```{r}
mean(data$bodytemp)

sd(data$bodytemp)
```

```{r}
data |>
  ggplot(aes(x = bodytemp)) +
  geom_histogram(aes(y = after_stat(density)),
                 col = "black", fill = "white") +
  geom_density(linewidth = 1) +
  theme_bw()
```

### Grid approximation of posterior given sample data


```{r}
mu = seq(96, 100, 0.01)
sigma = seq(0.01, 3, 0.01)

# theta is all possible (mu, sigma) pairs
theta = expand.grid(mu, sigma)
names(theta) = c("mu", "sigma")

# prior
prior_mu = dnorm(mu, 98.6, 0.3)
prior_sigma = dnorm(sigma, 0, 1)
prior = apply(expand.grid(prior_mu, prior_sigma), 1, prod)
prior = prior / sum(prior)

# data
y = data$bodytemp

# likelihood - looping through each theta, but probably better/easier ways
likelihood = rep(NA, nrow(theta))

for (i in 1:nrow(theta)) {
  likelihood[i] = prod(dnorm(y, theta[i, "mu"], theta[i, "sigma"]))
}

# posterior
product = likelihood * prior
posterior = product / sum(product)

```




```{r}
#| label: fig-body-temp-posterior-grid
#| echo: true
#| fig-cap: "Grid-approximated joint posterior distribution of mean and SD of body temperatures in @exm-body-temp-prior-sd."
#| fig-subcap: 
#|   - "Scatterplot."
#|   - "Density plot."
#| layout-ncol: 2

n_rep = 10000

# sample (mu, sigma) pairs for the posterior distribution by
# sampling rows of the Bayes table according to posterior
# and then finding the (mu, sigma) values for the sampled rows

index_sim = sample(1:length(posterior),
                   size = n_rep, replace = TRUE,
                   prob = posterior)

sim_posterior = theta |> slice(index_sim)

ggplot(sim_posterior,
       aes(x = mu,
           y = sigma)) +
  geom_point(color = bayes_col["posterior"], alpha = 0.4) +
  labs(x = "mean (mu)",
       y = "sd (sigma)") +
  theme_bw()

ggplot(sim_posterior,
       aes(x = mu,
           y = sigma)) +
  stat_density_2d(aes(fill = after_stat(level)),
                  geom = "polygon", color = "white") +
  scale_fill_viridis(option = "mako") +
  labs(x = "mean (mu)",
       y = "sd (sigma)") +
  theme_bw()
```

```{r}
cor(sim_posterior$mu, sim_posterior$sigma)
```

```{r}
mean(sim_posterior$mu)

sd(sim_posterior$mu)
```

```{r}
mean(sim_posterior$sigma)

sd(sim_posterior$sigma)
```

```{r}
#| label: fig-body-temp-posterior-grid-marginal
#| echo: true
#| fig-cap: "Grid-approximated marginal distribution of mean and SD of body temperatures in @exm-body-temp-prior-sd."
#| fig-subcap: 
#|   - "Population mean (mu)."
#|   - "Population SD (sigma)."
#| layout-ncol: 2


sim_posterior |>
  ggplot(aes(x = mu)) +
  geom_histogram(aes(y = after_stat(density)),
                 col = bayes_col["posterior"], fill = "white") +
  geom_density(linewidth = 1, col = bayes_col["posterior"]) +
  theme_bw()

sim_posterior |>
  ggplot(aes(x = sigma)) +
  geom_histogram(aes(y = after_stat(density)),
                 col = bayes_col["posterior"], fill = "white") +
  geom_density(linewidth = 1, col = bayes_col["posterior"]) +
  theme_bw()
```




### Posterior simulation with `brms`


```{r}
#| warning: false

library(brms)
library(bayesplot)
library(tidybayes)
```

Next we fit the model to the data using the function `brm`.
The three inputs are:

- Input 1: the data  
`data = data`

- Input 2: model for the data (likelihood)  
`family = gaussian`  
`bodytemp ~ 1 # linear model with intercept only`

- Input 3: prior distribution for parameters  
`prior(normal(98.6, 0.3), class = Intercept),`
`prior(normal(0, 1), class = sigma))`

- Other arguments specify the details of the numerical algorithm; our specifications below will result in 10000 $\theta$ values simulated from the posterior distribution.
(It will simulate 3500 values (`iter`), discard the first 1000 (`warmup`), and then repeat 4 times (`chains`).)


```{r}
#| cache: true

fit <- brm(data = data,
           family = gaussian,
           bodytemp ~ 1,
           prior = c(prior(normal(98.6, 0.3), class = Intercept),
                     prior(normal(0, 1), class = sigma)),
           iter = 3500,
           warmup = 1000,
           chains = 4,
           refresh = 0)
```


```{r}
summary(fit)
```

```{r}
plot(fit)
```



```{r}
posterior = fit |>
  spread_draws(b_Intercept, sigma) |>
  rename(mu = b_Intercept)
```


```{r}
posterior |> head(10) |> kbl() |> kable_styling()
```


Marginal posterior distribution of $\mu$

```{r}
#| warning: false

posterior |>
  ggplot(aes(x = mu)) +
  geom_histogram(aes(y = after_stat(density)),
                 col = bayes_col["posterior"], fill = "white") +
  geom_density(linewidth = 1, col = bayes_col["posterior"]) +
  theme_bw()
```

```{r}
# posterior mean of mu
mean(posterior$mu)

# posterior sd of mu
sd(posterior$mu)

# posterior quantiles are endpoints of credible intervals for mu
quantile(posterior$mu, c(0.01, 0.10, 0.25, 0.50, 0.75, 0.90, 0.99))
```

Marginal posterior distribution of $\sigma$

```{r}
#| warning: false

posterior |>
  ggplot(aes(x = sigma)) +
  geom_histogram(aes(y = after_stat(density)),
                 col = bayes_col["posterior"], fill = "white") +
  geom_density(linewidth = 1, col = bayes_col["posterior"]) +
  theme_bw()
```

```{r}
# posterior mean of sigma
mean(posterior$sigma)

# posterior sd of sigma
sd(posterior$sigma)

# posterior quantiles are endpoints of credible intervals for sigma
quantile(posterior$sigma, c(0.01, 0.10, 0.25, 0.50, 0.75, 0.90, 0.99))
```


Joint posterior distribution

```{r}
#| label: fig-body-temp-posterior-brms
#| echo: true
#| fig-cap: "Simulated joint posterior distribution of mean and SD of body temperatures in @exm-body-temp-brms-sd."
#| fig-subcap: 
#|   - "Scatterplot."
#|   - "Density plot."
#| layout-ncol: 2


ggplot(posterior,
       aes(x = mu,
           y = sigma)) +
  geom_point(color = bayes_col["posterior"], alpha = 0.4) +
  stat_ellipse(level = 0.98, color = "black", linewidth = 2) +
  stat_density_2d(color = "grey", linewidth = 1) +
  labs(x = "mean (mu)",
       y = "sd (sigma)") +
  theme_bw()

ggplot(posterior,
       aes(x = mu,
           y = sigma)) +
  stat_density_2d(aes(fill = after_stat(level)),
                  geom = "polygon", color = "white") +
  scale_fill_viridis(option = "mako") +
  labs(x = "mean (mu)",
       y = "sd (sigma)") +
  theme_bw()
```

```{r}
cor(posterior$mu, posterior$sigma)
```


### Posterior prediction


```{r}
n_rep = nrow(posterior)

posterior <- posterior |>
  mutate(y_sim = rnorm(n_rep,
                       mean = posterior$mu,
                       sd = posterior$sigma))

posterior |> head(10) |> kbl() |> kable_styling()
```


```{r}

ggplot(posterior,
       aes(x = y_sim)) +
  geom_histogram(aes(y = after_stat(density)),
                 col = bayes_col["posterior_predict"], fill = "white", bins = 100) +
  geom_density(linewidth = 1, col = bayes_col["posterior_predict"]) +
  labs(x = "Predicted body temp") +
  theme_bw()


# plot the cdf
ggplot(posterior,
       aes(x = y_sim)) + 
  stat_ecdf(col = bayes_col["prior_predict"],
            linewidth = 1) +
  labs(x = "Body temperature (y)",
       y = "cdf") +
  theme_bw()

```


```{r}
quantile(posterior$y_sim, c(0.025, 0.975))
```


### Posterior predictive checking


```{r}
pp_check(fit, type = "hist")
```


```{r}
pp_check(fit, ndraw = 100)
```


