# Log Linear Models for Counts

{{< include _r_setup.qmd >}}

{{< include _python_setup.qmd >}}




### Data

```{r}
n = 100

x = rnorm(n)

beta0 = rnorm(n, 0, 0.1)
beta1 = rnorm(n, 1, 0.1)

mu = exp(beta0 + beta1 * x)

y = rpois(n, mu)

data = data.frame(x, y)
```



### Posterior distribution


```{r}
library(brms)
library(tidybayes)
library(bayesplot)
```

```{r}
get_prior(data = data,
          family = poisson(),
          y ~ 1 + x)
```



```{r}
fit <- brm(data = data,
           family = poisson(),
           y ~ 1 + x,
           prior = c(prior(normal(4, 0.1), class = Intercept),
                     prior(normal(0, 0.1), class = b)),
           iter = 3500,
           warmup = 1000,
           chains = 4,
           refresh = 0)
```


```{r}
summary(fit)
```


```{r}
plot(fit)
```


```{r}
pairs(fit)
```


```{r}
mcmc_dens_overlay(fit, pars = vars(b_Intercept, b_BMI, sigma))
```



```{r}
neff_ratio(fit)[c("b_Intercept", "b_BMI", "sigma")]
```

```{r}
neff_ratio(fit)[c("b_Intercept", "b_BMI", "sigma")] |> 
  mcmc_neff() +
  yaxis_text(hjust = 0) 
```


### Posterior distribution, model using centered BMI

```{r}
data = data |>
  mutate(BMI_c = BMI - mean(BMI))
```


```{r}
fit_c <- brm(data = data,
           family = gaussian(),
           PBF ~ 1 + BMI_c,
           prior = c(prior(normal(18, 2), class = Intercept),
                prior(normal(0.5, 0.1), class = b),
                prior(exponential(0.667), class = sigma)),
           iter = 3500,
           warmup = 1000,
           chains = 4,
           refresh = 0)
```


```{r}
summary(fit_c)
```


```{r}
color_scheme_set("viridis")

plot(fit_c)
```


```{r}
pairs(fit_c)
```



```{r}
neff_ratio(fit_c)[c("b_Intercept", "b_BMI_c", "sigma")]
```


```{r}
neff_ratio(fit_c)[c("b_Intercept", "b_BMI_c", "sigma")] |> 
  mcmc_neff() +
  yaxis_text(hjust = 0) 
```


Posterior distribution of regression slope

```{r}
mcmc_intervals(fit_c,
               pars = c("b_BMI_c"),
               prob = 0.5,
               prob_outer = 0.98)
```

```{r}
#| layout-nrow: 3

color_scheme_set("teal")

mcmc_areas(fit_c,
           pars = c("b_BMI_c"),
           prob = 0.5,
           point_est = "median")

mcmc_areas(fit_c,
           pars = c("b_BMI_c"),
           prob = 0.8,
           point_est = "median")

mcmc_areas(fit_c,
           pars = c("b_BMI_c"),
           prob = 0.98,
           point_est = "median")
```




Posterior distribution of $\sigma$

```{r}
mcmc_intervals(fit_c,
               pars = c("sigma"),
               prob = 0.5,
               prob_outer = 0.98)
```

```{r}
#| layout-nrow: 3

mcmc_areas(fit_c,
           pars = c("sigma"),
           prob = 0.5,
           point_est = "median")

mcmc_areas(fit_c,
           pars = c("sigma"),
           prob = 0.8,
           point_est = "median")

mcmc_areas(fit_c,
           pars = c("sigma"),
           prob = 0.98,
           point_est = "median")
```



Posterior distribution of mean response

```{r}
posterior_c = fit_c |>
  spread_draws(b_Intercept, b_BMI_c, sigma) |>
  mutate(mu_20 = b_Intercept + b_BMI_c * (20 - mean(data$BMI)))

posterior_c |> head(10) |> kbl() |> kable_styling()
```


```{r}
posterior_c |>
  ggplot(aes(x = mu_20)) +
  stat_halfeye(.width = c(0.80, 0.98),
               fill = bayes_col["posterior"]) +
  theme_bw()
```



```{r}
quantile(posterior_c$mu_20, c(0.01, 0.99))
```



### Posterior prediction


```{r}
posterior_c = posterior_c |>
  mutate(y_pred_20 = mu_20 + sigma * rnorm(nrow(posterior_c), 0, 1))

posterior_c |> head(10) |> kbl() |> kable_styling()
```


```{r}
ggplot(posterior_c,
       aes(x = y_pred_20)) +
  geom_histogram(aes(y = after_stat(density)),
                 col = bayes_col["posterior_predict"], fill = "white", bins = 100) +
  geom_density(linewidth = 1, col = bayes_col["posterior_predict"]) +
  labs(x = "Posterior predicted PBF(%) for BMI of 20 kg/m^2") +
  theme_bw()
```

```{r}
quantile(posterior_c$y_pred_20, c(0.025, 0.975))
```



```{r}
predictive_interval(fit_c,
                    newdata = data.frame(BMI_c = 20 - mean(data$BMI)),
                    prob = 0.95)
```


### Posterior predictive checking

4 simulated samples

```{r}
data |>
  add_predicted_draws(fit_c, ndraws = 4) %>%
  ggplot(aes(x = BMI_c, y = PBF)) +
    geom_point(aes(y = .prediction,
                   group = .draw), size = 0.2) + 
    facet_wrap(~ .draw)
```

Posterior predictive check of response

```{r}
color_scheme_set("purple")

pp_check(fit_c, ndraw = 100)

```

`posterior_predict` simulates a sample of size 2157, using the observed $x$ values, for each of the 10000 simulated triples from the posterior distribution, resulting in a 10000 x 2157 matrix.

```{r}
predictions <- posterior_predict(fit_c)

dim(predictions)
```


We can use these predictions to plot posterior prediction intervals for a range of BMI values.

```{r}
ppc_intervals(data$PBF,
              yrep = predictions,
              x = data$BMI, 
              prob = 0.5,
              prob_outer = 0.95)
```





### Sensitivity: using `brms` default prior



```{r}
get_prior(data = data,
          family = gaussian(),
          PBF ~ 1 + BMI_c)
```


```{r}
fit_brms_prior <- brm(data = data,
           family = gaussian(),
           PBF ~ 1 + BMI_c,
           sample_prior = TRUE,
           iter = 3500,
           warmup = 1000,
           chains = 4,
           refresh = 0)
```


```{r}
summary(fit_brms_prior)
```


```{r}
color_scheme_set("viridis")

plot(fit_brms_prior)
```


```{r}
pairs(fit_brms_prior)
```



```{r}
neff_ratio(fit_brms_prior)[c("b_Intercept", "b_BMI_c", "sigma")]
```


```{r}
neff_ratio(fit_brms_prior)[c("b_Intercept", "b_BMI_c", "sigma")] |> 
  mcmc_neff() +
  yaxis_text(hjust = 0) 
```


```{r}
color_scheme_set("purple")

pp_check(fit_brms_prior, ndraw = 100)

```

```{r}
predictions_b <- posterior_predict(fit_brms_prior)

dim(predictions_b)
```




```{r}
ppc_intervals(data$PBF,
              yrep = predictions_b,
              x = data$BMI, 
              prob = 0.5,
              prob_outer = 0.95)
```

Leave-one-out (LOO) cross-validation to compare models.

```{r}
LOO(fit_c, fit_brms_prior)
```





