# Bayesian Simple Linear Regression Model {#sec-regression}

{{< include _r_setup.qmd >}}

<!-- {{< include _python_setup.qmd >}} -->







Regression is one of the most widely used statistical techniques for modeling relationships between variables.
We will now consider a Bayesian treatment of *simple linear regression*.
We'll use the following example throughout.

[Percent body fat](https://en.wikipedia.org/wiki/Body_fat_percentage) (PBF, total mass of fat divided by total body mass) is an indicator of physical fitness level, but it is difficult to measure accurately.
Body mass index (BMI), which is easily computed based on height and weight ($\text{BMI} = \frac{\text{weight in kg}}{\text{(height in m)}^2}$), is another commonly used measure^[But BMI has its own problems; see <http://fivethirtyeight.com/features/bmi-is-a-terrible-measure-of-health/>.].
What is the relationship between BMI and percent body fat?
How can we use a person's BMI to estimate their percent body fat?

We'll fit a Bayesian model for predicting percent body fat from BMI using data  on a nationally representative sample of 2157 U.S. males from the [National Health and Nutrition Examination Survey (NHANES)](http://www.cdc.gov/nchs/nhanes/about_nhanes.htm).






::: {#exm-regression-prior}
Consider the assumptions of a Bayesian simple linear regression model for predicting percent body fat from BMI.
:::


1. What are the assumptions of the simple linear regression model?
(There are many different regression models.
We're referring to the one that is usually the first one you see in introductory statistics.
Sometimes the assumptions are summarized with the acronym "LINE".)
\
\
\
\
\

1. What are the parameters in the model?
That is, what do we need to put a prior distribution on (in order to eventually get a posterior distribution given sample data)?
What do the parameters represent in context?
\
\
\
\
\

1. Suggest a prior distribution.
Explain your reasoning.
\
\
\
\
\

1. Describe in full detail how you would simulate the prior predictive distribution of percent body fat for males with a BMI of 20.
\
\
\
\
\
\

1. Simulate the prior predictive distribution of percent body fat for a range of BMI values.
Do the distributions seem reasonable?
If not, experiment with different priors until you find one that produces a reasonable prior predictive distribution.
\
\
\
\
\





- Consider data on a numerical response variable $y$ and a numerical explanatory (or predictor) variable $x$.
A **simple linear regression model** of $y$ on $x$ assumes:
    -  *Linearity*: the conditional mean of $y$ is *linear* in $x$; for some slope $\beta_1$ and intercept $\beta_0$:
    $$
    \text{E}(y|x) = \beta_0 + \beta_1 x
    $$
    - *Independence*: $y_1,\ldots, y_n$ are independent^[In a Bayesian context, we assume conditional independence of $y_1, \ldots, y_n$ given the parameters $\beta_0, \beta_1, \sigma$]
    - *Normality*: for each $x$, the conditional distribution of $y$ given $x$ is Normal
    - *Equal variance*: the conditional variance of $y$ does not depend on $x$
    $$
    \text{Var}(y|x) = \sigma^2
    $$
- The regression model is *conditional*.
That is, we make assumptions about and perform inference for the conditional distribution of the response variable $y$ given values of the explanatory variable $x$.
Therefore, even when we technically have a sample of pairs $(x_i,y_i)$ from a bivariate distribution, we will always condition on the values of $x$ and treat them as non-random.
- Linearity is an assumption: we assume there is some "true" population *regression line* $\beta_0 + \beta_1 x$ which specifies the conditional *mean* of the response variable for each value of the explanatory variable.
- The above set of assumptions can be stated more compactly as
$$
y_i \sim N\left( \beta_0 + \beta_1 x_i, \sigma^2\right), \text{ and } y_1, \ldots, y_n \text{ are independent}
$$
Equivalently,
$$
y_i = \beta_0 + \beta_1 x_i + \varepsilon_i, \text{ where } \varepsilon_i, \ldots, \varepsilon_n \text{ are i.i.d. } N(0, \sigma).
$$
- The parameters of the above model are $\beta_0$, $\beta_1$, and $\sigma$.
- $\sigma_y^2$ denotes the variance of the response variable $y$ with respect to its overall mean, while $\sigma^2$ denotes the variance of values of the response variable  about the true regression line.
Thus, $\sigma^2 = (1-\rho^2)\sigma^2_y$, where $\rho$ is the population correlation between the explanatory and response variables.


::: {#exm-regression-freq}
@sec-regression-sample summarizes the sample data.
Before fitting a Bayesian model, we'll examine the usual frequentist results.
:::





1. Find and interpret the slope of the fitted regression line.
\
\
\
\
\

1. Find and interpret a 95% confidence interval for the slope of the regression line.
(How do you interpret 95% confidence?)
\
\
\
\
\

1. Find the equation of the fitted regression line.
\
\
\
\
\

1. Find and interpret the fitted value for men with a BMI of 20 kg/m^2^.
\
\
\
\
\

1. Find and interpret a 95% confidence interval based on the fitted value for men with a BMI of 20 kg/m^2^.
\
\
\
\
\

1. Find and interpret a point estimate of $\sigma$.
\
\
\
\
\
\

1. Find and interpret a 95% *prediction* interval for men with a BMI of 20 kg/m^2^.
\
\
\
\
\




::: {#exm-regression-posterior}
Now we'll fit a Bayesian model and find the posterior distribution.
:::



1. Suppose there are just two observations of (BMI, PBF): (15, 20) and (40, 30).
How would you compute the likelihood?
\
\
\
\
\

1. Use `brms` to fit the model to the full sample data.
Run some diagnostic checks; does the MCMC algorithm seem to be working efficiently?
\
\
\
\
\

1. Inspect the joint posterior distribution of $\beta_0$ and $\beta_1$.
What do you notice?
Why do you think this happens?
\
\
\
\
\

1. Now suppose we replace BMI ($x$) with its centered version ($x - \bar{x}$).
What does the parameter $\beta_0$ represent in this model?
Suggest a prior distribution for $\beta_0$ (considering it in isolation).
\
\
\
\
\

1. Use `brms` to fit the model using centered BMI as the explanatory variable.
Inspect the joint posterior distribution of $\beta_0$ and $\beta_1$.
What do you notice?
Why do you think this happens?
\
\
\
\
\



::: {#exm-regression-conclusions}
Now we'll summarize and report conclusions based on the posterior distribution, using the model with centered BMI.
:::


1. Describe the posterior distribution of $\beta_1$.
Find and interpret 50%, 80%, and 98% posterior credible intervals for $\beta_1$.
\
\
\
\
\

1. What does the parameter $\mu_{20} = \beta_0 + \beta_1(20 -\bar{x})$ represent in context?
Explain how you could approximate the posterior distribution of $\mu_{20}$.
Then approximate the posterior distribution, describe it, and
find and interpret a 98% posterior credible interval for $\mu_{20}$.
\
\
\
\
\

1. Describe the posterior distribution of $\sigma$.
Find and interpret 50%, 80%, and 98% posterior credible intervals for $\sigma$.
\
\
\
\
\

1. Compare the results of the Bayesian and frequentist analyses. 
How are they similar?
How are they different?
\
\
\
\
\

::: {#exm-regression-prediction}
Now we'll use the fitted model to predict percent body fat.
:::

1. Explain how you could approximate the posterior predictive distribution of percent body fat for men with a BMI of 20 kg/m^2^.
\
\
\
\
\

1. Find and interpret a 95% posterior prediction interval for men with a BMI of 20 kg/m^2^.
\
\
\
\
\


1. Compare the 95% posterior prediction interval with the corresponding frequentist 95% prediction interval.
How are they similar?
How are they different?
\
\
\
\
\


1. Suggest how you could perform a posterior predictive check.
Then use software to perform the check.
Does the model seem reasonable based on the data?
\
\
\
\
\



- There are a number of ways the assumptions of the simple linear regression model could be revised.
- We could assume the mean of $y$ changes with $x$ is a non-linear way (e.g., quadratic).
- We could assume distributions other than Normal in the likelihood.
For example, we could assume a $t$-distribution if outliers were an issue.
- We could assume the conditional variance of $y$ also changes with $x$.
- We could include multiple explanatory variables $x$.
- Each of the above would be a change in the *likelihood*, perhaps introducing additional parameters.


::: {#exm-regression-brms}
Now we'll investigate sensitivity of the results to choice of prior.
Rerun the model (using centered BMI) but let `brms` choose the prior.
Compare the results of the two Bayesian models; are the results very sensitive to the choice of prior?
:::
\
\
\
\
\



## Notes


### Prior predictive tuning


- We'll start with a prior distribution under which $\beta_0$, $\beta_1$, and $\sigma$ are independent.
- A quick Google search reveals that for men [percent body fat ranges from about 6% to 30%](https://www.webmd.com/diet/features/body-fat-measurement), and [BMI ranges from about 15 to 35 kg/m^2^](https://www.cdc.gov/healthyweight/assessing/bmi/adult_bmi/index.html).
- The overall range of percent body fat is about 6% to 30%, but we would expect less variability among men of the same BMI.
If we assume percent body fat varies by about 5 percentage points among men of the same BMI, we can choose a prior for $\sigma$ that has a mean of 5, say Exponential(1/5) (to reflect $\sigma>0$).
- We expect a positive association between percent body fat and BMI, so we want a prior for $\beta_1$ that places most of its credibility on positive values.
Disregarding measurement units, the two variables take values on fairly similar numerical scales (6 to 30, 15 to 35).
For the slope $\beta_1$, we might assume a Normal prior with a mean of 1 and a standard deviation of 0.5 (to keep most of the credibility on positive values, reflecting prior belief in positive association).
- The intercept $\beta_0$, in principle, represents the percent body fat for a man with a BMI of 0, so  we might assume a Normal prior with a mean of 0 and a standard deviation of 1 (to put most of the prior credibility near 0).
(We'll see another way to think about this parameter soon.)
In summary, our prior distribution on the parameters $(\beta_0, \beta_1, \sigma)$ assumes
    - $\beta_0$, $\beta_1$, and $\sigma$ are independent
    - $\sigma \sim$ Exponential(1/5)
    - $\beta_1\sim$ Normal(1, 0.5)
    - $\beta_0\sim$ Normal(0, 1)

Prior predictive distribution for BMI of 20

```{r}
n_rep = 10000

# value of x
x = 20

# simulate parameters from prior distribution
beta0 = rnorm(n_rep, 0, 1)
beta1 = rnorm(n_rep, 1, 0.5)
sigma = rexp(n_rep, 1 / 5)

# simulate values of percent body fat based on assumptions of regression model
y = rnorm(n_rep, beta0 + beta1 * x, sigma)

hist(y, xlab = "Prior predicted PBF(%) for BMI of 20 kg/m^2",
     breaks = 100,
     col = bayes_col["prior_predict"])
```


Prior predictive distribution for many values of BMI

```{r}
n_rep = 1000

# a range of values of x
x = runif(n_rep, 15, 35)

# simulate parameters from prior distribution
beta0 = rnorm(n_rep, 0, 1)
beta1 = rnorm(n_rep, 1, 0.5)
sigma = rexp(n_rep, 1 / 5)

# simulate values of percent body fat based on assumptions of regression model
y = rnorm(n_rep, beta0 + beta1 * x, sigma)

# scatter plot or prior predicted values
plot(x, y,
     xlab = "BMI (kg/m^2)", ylab = "Prior predicted percent body fat (%)",
     col = bayes_col["prior_predict"])
```

What regression lines might look like according to prior distribution

```{r}
# number of lines to plot
n_lines = 100

# blank scatterplot - just set the plot area and scale
plot(x, y, xlab = "BMI (kg/m^2)", ylab = "Percent body fat (%)", type = "n",
 xlim = c(15, 35),
 ylim = c(0, 30))

# plot each of the simulated lines
# using slope and intercept simulated from prior distribution
for (l in 1:n_lines){
  abline(beta0[l], beta1[l], col = bayes_col["prior"], lwd = 1)
}
```

### Prior predictive distribution

After some tuning, our prior distribution on the parameters $(\beta_0, \beta_1, \sigma)$ assumes

- $\beta_0$, $\beta_1$, and $\sigma$ are independent
- $\sigma \sim$ Exponential(1/1.5)
- $\beta_1\sim$ Normal(0.5, 0.1)
- $\beta_0\sim$ Normal(2, 2)

Prior predictive distributions

```{r}
n_rep = 1000

# a range of values of x
x = runif(n_rep, 15, 35)

# simulate parameters from prior distribution
beta0 = rnorm(n_rep, 2, 2)
beta1 = rnorm(n_rep, 0.5, 0.1)
sigma = rexp(n_rep, 1 / 1.5)

# simulate values of percent body fat based on assumptions of regression model
y = rnorm(n_rep, beta0 + beta1 * x, sigma)

# scatter plot or prior predicted values
plot(x, y,
     xlab = "BMI (kg/m^2)", ylab = "Prior predicted percent body fat (%)",
     col = bayes_col["prior_predict"])
```


```{r}
# number of lines to plot
n_lines = 100

# blank scatterplot - just set the plot area and scale
plot(x, y, xlab = "BMI (kg/m^2)", ylab = "Percent body fat (%)", type = "n",
 xlim = c(15, 35),
 ylim = c(0, 30))

# plot each of the simulated lines
# using slope and intercept simulated from prior distribution
for (l in 1:n_lines){
  abline(beta0[l], beta1[l], col = bayes_col["prior"], lwd = 1)
}
```


### Sample data {#sec-regression-sample}

```{r}
data = read_csv("nhanes_males.csv")

data |> head() |> kbl() |> kable_styling()
```

```{r}
c(summary(data$BMI), SD = sd(data$BMI))
```

```{r}
c(summary(data$PBF), SD = sd(data$PBF))
```

```{r}
cor(data$BMI, data$PBF)
```


```{r}
data_plot = ggplot(data, 
                   aes(x = BMI, y = PBF)) +
  geom_point(shape = 1, size = 2) +
  labs(x = "BMI (kg/m^2)",
       y = "Percent body fat (%)") +
  theme_bw()

data_plot
```


### Frequentist results

```{r}
freq = lm(PBF ~ BMI, data)

summary(freq)
```

Confidence interval for mean response

```{r}
x_new = data.frame(BMI = 20)

predict(freq, newdata = x_new, interval = "confidence", se.fit = TRUE)
```

```{r}
#| warning: false

data_plot +
  geom_smooth(method = lm,
              col = bayes_col["likelihood"],
              fill = bayes_col["likelihood"])
```

Prediction intervals

```{r}
x_new = data.frame(BMI = 20)

predict(freq, newdata = x_new, interval = "prediction")
```

```{r}
#| warning: false

predictions <- predict(freq, interval = "predict")

data |>
  cbind(predictions) |>
  ggplot(aes(x = BMI, y = PBF)) +
  geom_point(shape = 1, size = 2) +
  geom_ribbon(aes(ymin = lwr, ymax = upr),
              fill = bayes_col["likelihood"], alpha = 0.4) +
  labs(x = "BMI (kg/m^2)",
       y = "Percent body fat (%)") +
  theme_bw()


```




### Posterior distribution


```{r}
library(brms)
library(tidybayes)
library(bayesplot)
```



```{r}
fit <- brm(data = data,
           family = gaussian(),
           PBF ~ 1 + BMI,
           prior = c(prior(normal(2, 2), class = Intercept),
                     prior(normal(0.5, 0.1), class = b),
                     prior(exponential(0.667), class = sigma)),
           iter = 3500,
           warmup = 1000,
           chains = 4,
           refresh = 0)
```


```{r}
summary(fit)
```


```{r}
plot(fit)
```


```{r}
pairs(fit)
```


```{r}
mcmc_dens_overlay(fit, pars = vars(b_Intercept, b_BMI, sigma))
```



```{r}
neff_ratio(fit)[c("b_Intercept", "b_BMI", "sigma")]
```

```{r}
neff_ratio(fit)[c("b_Intercept", "b_BMI", "sigma")] |> 
  mcmc_neff() +
  yaxis_text(hjust = 0) 
```


### Posterior distribution, model using centered BMI

```{r}
data = data |>
  mutate(BMI_c = BMI - mean(BMI))
```


```{r}
fit_c <- brm(data = data,
           family = gaussian(),
           PBF ~ 1 + BMI_c,
           prior = c(prior(normal(18, 2), class = Intercept),
                prior(normal(0.5, 0.1), class = b),
                prior(exponential(0.667), class = sigma)),
           iter = 3500,
           warmup = 1000,
           chains = 4,
           refresh = 0)
```


```{r}
summary(fit_c)
```


```{r}
color_scheme_set("viridis")

plot(fit_c)
```


```{r}
pairs(fit_c)
```



```{r}
neff_ratio(fit_c)[c("b_Intercept", "b_BMI_c", "sigma")]
```


```{r}
neff_ratio(fit_c)[c("b_Intercept", "b_BMI_c", "sigma")] |> 
  mcmc_neff() +
  yaxis_text(hjust = 0) 
```


Posterior distribution of regression slope

```{r}
mcmc_intervals(fit_c,
               pars = c("b_BMI_c"),
               prob = 0.5,
               prob_outer = 0.98)
```

```{r}
#| layout-nrow: 3

color_scheme_set("teal")

mcmc_areas(fit_c,
           pars = c("b_BMI_c"),
           prob = 0.5,
           point_est = "median")

mcmc_areas(fit_c,
           pars = c("b_BMI_c"),
           prob = 0.8,
           point_est = "median")

mcmc_areas(fit_c,
           pars = c("b_BMI_c"),
           prob = 0.98,
           point_est = "median")
```




Posterior distribution of $\sigma$

```{r}
mcmc_intervals(fit_c,
               pars = c("sigma"),
               prob = 0.5,
               prob_outer = 0.98)
```

```{r}
#| layout-nrow: 3

mcmc_areas(fit_c,
           pars = c("sigma"),
           prob = 0.5,
           point_est = "median")

mcmc_areas(fit_c,
           pars = c("sigma"),
           prob = 0.8,
           point_est = "median")

mcmc_areas(fit_c,
           pars = c("sigma"),
           prob = 0.98,
           point_est = "median")
```



Posterior distribution of mean response

```{r}
posterior_c = fit_c |>
  spread_draws(b_Intercept, b_BMI_c, sigma) |>
  mutate(mu_20 = b_Intercept + b_BMI_c * (20 - mean(data$BMI)))

posterior_c |> head(10) |> kbl() |> kable_styling()
```


```{r}
posterior_c |>
  ggplot(aes(x = mu_20)) +
  stat_halfeye(.width = c(0.80, 0.98),
               fill = bayes_col["posterior"]) +
  theme_bw()
```



```{r}
quantile(posterior_c$mu_20, c(0.01, 0.99))
```



### Posterior prediction


```{r}
posterior_c = posterior_c |>
  mutate(y_pred_20 = mu_20 + sigma * rnorm(nrow(posterior_c), 0, 1))

posterior_c |> head(10) |> kbl() |> kable_styling()
```


```{r}
ggplot(posterior_c,
       aes(x = y_pred_20)) +
  geom_histogram(aes(y = after_stat(density)),
                 col = bayes_col["posterior_predict"], fill = "white", bins = 100) +
  geom_density(linewidth = 1, col = bayes_col["posterior_predict"]) +
  labs(x = "Posterior predicted PBF(%) for BMI of 20 kg/m^2") +
  theme_bw()
```

```{r}
quantile(posterior_c$y_pred_20, c(0.025, 0.975))
```



```{r}
predictive_interval(fit_c,
                    newdata = data.frame(BMI_c = 20 - mean(data$BMI)),
                    prob = 0.95)
```


### Posterior predictive checking

4 simulated samples

```{r}
data |>
  add_predicted_draws(fit_c, ndraws = 4) %>%
  ggplot(aes(x = BMI_c, y = PBF)) +
    geom_point(aes(y = .prediction,
                   group = .draw), size = 0.2) + 
    facet_wrap(~ .draw)
```

Posterior predictive check of response

```{r}
color_scheme_set("purple")

pp_check(fit_c, ndraw = 100)

```

`posterior_predict` simulates a sample of size 2157, using the observed $x$ values, for each of the 10000 simulated triples from the posterior distribution, resulting in a 10000 x 2157 matrix.

```{r}
predictions <- posterior_predict(fit_c)

dim(predictions)
```


We can use these predictions to plot posterior prediction intervals for a range of BMI values.

```{r}
ppc_intervals(data$PBF,
              yrep = predictions,
              x = data$BMI, 
              prob = 0.5,
              prob_outer = 0.95)
```





### Sensitivity: using `brms` default prior



```{r}
get_prior(data = data,
          family = gaussian(),
          PBF ~ 1 + BMI_c)
```


```{r}
fit_brms_prior <- brm(data = data,
           family = gaussian(),
           PBF ~ 1 + BMI_c,
           sample_prior = TRUE,
           iter = 3500,
           warmup = 1000,
           chains = 4,
           refresh = 0)
```


```{r}
summary(fit_brms_prior)
```


```{r}
color_scheme_set("viridis")

plot(fit_brms_prior)
```


```{r}
pairs(fit_brms_prior)
```



```{r}
neff_ratio(fit_brms_prior)[c("b_Intercept", "b_BMI_c", "sigma")]
```


```{r}
neff_ratio(fit_brms_prior)[c("b_Intercept", "b_BMI_c", "sigma")] |> 
  mcmc_neff() +
  yaxis_text(hjust = 0) 
```


```{r}
color_scheme_set("purple")

pp_check(fit_brms_prior, ndraw = 100)

```

```{r}
predictions_b <- posterior_predict(fit_brms_prior)

dim(predictions_b)
```




```{r}
ppc_intervals(data$PBF,
              yrep = predictions_b,
              x = data$BMI, 
              prob = 0.5,
              prob_outer = 0.95)
```

Leave-one-out (LOO) cross-validation to compare models.

```{r}
LOO(fit_c, fit_brms_prior)
```





